---
title: "GE_Gravity Logic and Testing"
author: "Tom Zylkin, Vadim Kudlay"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GE_Gravity Theory, Logic, Testing, and Comparison}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, echo=FALSE}
library(GEGravity)
```

So, we just discussed all of the logic behind it, so let us just run the example normally and get the data.

# Running The Algorithm 

```{r, echo=FALSE}
library(alpaca)  # Needed for partial coefficients
```

#### Pre-Processing
```{r}
# Foreign trade subset
f_trade <- TradeData0014[TradeData0014$exporter != TradeData0014$importer,]
```

```{r}
# Normalize trade data to unit interval
f_trade$trade <- f_trade$trade / max(f_trade$trade)

# classify FEs for components to be absorbed (finding variable interactions)
f_trade$exp_year <- interaction(f_trade$expcode, f_trade$year)
f_trade$imp_year <- interaction(f_trade$impcode, f_trade$year)
f_trade$pair     <- interaction(f_trade$impcode, f_trade$expcode)

# Fit generalized linear model based on specifications
partials <- feglm(
  formula = trade ~ eu_enlargement + other_fta | exp_year + imp_year + pair, 
  data    = f_trade, 
  family  = poisson()
)$coefficient  # We just need the coefficients for computation
```

```{r}
# Sort trade matrix to make it easier to find imp/exp pairs
t_trade <- TradeData0014[order(
  TradeData0014$exporter, 
  TradeData0014$importer, 
  TradeData0014$year
),]

t_trade$eu_effect <- NA      # This creates new column eu_effect

i <- 1
# Effect of EU entrance on country based on partial, if entry happened
invisible(by(t_trade, list(t_trade$expcode, t_trade$impcode), function(row) { 
  
  # This is an effective boolean; was this a new EU pair created within time span?
  is_new_eu_pair <- tail(row$eu_enlargement, 1) - head(row$eu_enlargement, 1)

  # If it was, give it the computed partial eu_enlargement coefficient; if not, leave it unset
  t_trade[i, "eu_effect"] <<- is_new_eu_pair * partials[1]
  
  i <<- i + nrow(row)         # Increment index to start at the next imp/exp pair
  
}))

# Data to be finally fed to the function; only subset with eu_effect will be used
data <- t_trade[t_trade$year == 2000,]   # In example, 1892 Entries, 5676 removed
head(data)
```

#### Running Actual Computations

```{r}
w_mult = ge_gravity(
  exp_id = data$expcode,    # Origin country associated with each observation
  imp_id = data$impcode,    # Destination country associated with each observation
  flows  = data$trade,      # Observed trade flows in the data for the year being used as the baseline
  beta   = data$eu_effect,  # “Partial” change in trade, obtained as coefficient from gravity estimation
  theta  = 4,               # Trade elasticity
  mult   = TRUE,            # 
  data   = data
)

w_o_mult = ge_gravity(
  exp_id = data$expcode,    # Origin country associated with each observation
  imp_id = data$impcode,    # Destination country associated with each observation
  flows  = data$trade,      # Observed trade flows in the data for the year being used as the baseline
  beta   = data$eu_effect,  # “Partial” change in trade, obtained as coefficient from gravity estimation
  theta  = 4,                # Trade elasticity
  mult   = FALSE,           # Assume trade balance is an additive component of national expenditure
  data   = data
)
```

#### Final results without multiplication parameter

```{r}
head(w_o_mult)
```

#### Final results with multiplication parameter

```{r}
head(w_mult)
```

***

# Comparison with Stata Counterpart

As mentioned, this package is intended to mimic the functionality of the Stata package of the same name, so we will do a quick comparison of this data relative to the same computation from the Stata counterpart. 

Before running comparisons, we need to slightly modify the results data to sync with our new format.

```{r}
# Notice that the Stata counterpart returned all years with a sparse 
# selection labeled with computed values. Ours just returns the new data 
# by default or tags it onto the data provided in the `data` parameter.

# To make it sync, just extract a year.
results <- TradeData0014_Results[TradeData0014_Results$year == 2000, ]
head(results)
```

#### Comparison of w_eu from results to the welfare w/o multiplier
```{r}
plot(x = results$w_eu, y = w_o_mult$welfare, log = "xy")
abline(coef = c(0,1))

message("Max difference: ", max(abs(results$w_eu - w_o_mult$welfare)))
```

#### Comparison of w_mult from results to the welfare with multiplier
```{r}
plot(x = results$w_mult, y = w_mult$welfare, log = "xy")
abline(coef = c(0,1))

message("Max difference: ", max(abs(results$w_mult - w_mult$welfare)))
```

#### Comparing results of new X w/o multiplier option
```{r}
plot(x = results$X_eu, y = w_o_mult$new_trade, log = "xy")
abline(coef = c(0,1))

message("Max difference: ", max(abs(results$X_eu - w_o_mult$new_trade)))
```

#### Comparing results of new X with multiplier option
```{r}
plot(x = results$X_mult, y = w_mult$new_trade, log = "xy")
abline(coef = c(0,1))

message("Max difference: ", max(abs(results$X_mult - w_mult$new_trade)))
```
